Substructure Boosts to Dark Matter Annihilation from Sommerfeld Enhancement

### 2009-02-13 tex/somm.tex
Updated draft

\newcommand{\eg}{e.g.}
\newcommand{\Fermi}{\emph{Fermi}}
\newcommand{\VL}{Via Lactea}
\newcommand{\aquarius}{Aquarius}
\newcommand{\CDM}{CDM}
\newcommand{\LCDM}{\ensuremath{\Lambda}\CDM}
\newcommand{\NFW}{NFW}
\newcommand{\GNFW}{G\NFW}
\newcommand{\DM}{DM}
\newcommand{\SM}{SM}
\newcommand{\OM}{OM}
\newcommand{\mpl}{\ensuremath{m_{\mathrm{Pl}}}}
\newcommand{\somm}{\ensuremath{S}}
\newcommand{\mdm}{\ensuremath{m_{\chi}}}
\newcommand{\mv}{\ensuremath{m_{\phi}}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\Eqnname}{Equation}
\newcommand{\eqnname}{equation}
\newcommand{\Ngamma}{\ensuremath{N_{\gamma}}}
\newcommand{\Ngammai}{\ensuremath{N_{\gamma,i}}}
\newcommand{\Eth}{\ensuremath{E_{\mathrm{th}}}}
\newcommand{\sigmaannv}{\ensuremath{\langle\sigma v\rangle}}
\newcommand{\los}{los}
\newcommand{\order}{\ensuremath{\mathcal{O}}}
\newcommand{\sigv}{\ensuremath{\sigma_v}}
\newcommand{\lum}{\ensuremath{\mathcal{L}}}
\newcommand{\lumsmooth}{\ensuremath{\lum_{\mbox{{\footnotesize sm}}}}}
\newcommand{\lumsmoothc}{\ensuremath{\lum_{\mbox{{\footnotesize sm}},v=c}}}
\newcommand{\lumsmoothtoo}{\ensuremath{\overline{\mathcal{L}}}}
\newcommand{\lumsub}{\ensuremath{\lum_{\mbox{{\footnotesize sub}}}}}
\newcommand{\boost}{\ensuremath{B}}
\newcommand{\redr}{\ensuremath{\tilde{r}}}
\newcommand{\rhos}{\ensuremath{\rho_s}}
\newcommand{\rs}{\ensuremath{r_s}}
\newcommand{\dist}{\ensuremath{D}}
\newcommand{\norm}{\ensuremath{\mathcal{N}}}
\newcommand{\normnfw}{\ensuremath{\norm_1}}
\newcommand{\normeinasto}{\ensuremath{\norm_2}}
\newcommand{\rhominustwo}{\ensuremath{\rho_{-2}}}
\newcommand{\rminustwo}{\ensuremath{r_{-2}}}
\newcommand{\sommboost}{\ensuremath{\mathcal{S}}}
\newcommand{\clumpboost}{\ensuremath{\mathcal{C}}}
\newcommand{\alphaEinasto}{\ensuremath{\alpha}}
\newcommand{\Msol}{\ensuremath{M_{\odot}}}
\newcommand{\Mvir}{\ensuremath{M_{\mbox{{\footnotesize vir}}}}}
\newcommand{\Rvir}{\ensuremath{r_{\mbox{{\footnotesize vir}}}}}
\newcommand{\conc}{\ensuremath{c}}
\newcommand{\rhoc}{\ensuremath{\rho_{\mbox{{\footnotesize c}}}}}
\newcommand{\omegam}{\ensuremath{\Omega_{m}}}
\newcommand{\vcirc}{\ensuremath{v_{\mbox{{\footnotesize c}}}}}
\newcommand{\vmax}{\ensuremath{v_{\mbox{{\footnotesize max}}}}}
\newcommand{\rmax}{\ensuremath{r_{\mbox{{\footnotesize max}}}}}

%Journals
\newcommand{\aap}{Astron.~Astrophys.}
\renewcommand{\nat}{Nature}
\newcommand{\jhep}{J.~High.~Energy.~Phys.}
\newcommand{\astropartphys}{Astropart.~Phys.}
\newcommand{\mnras}{Mon.~Not.~Roy.~Astron.~Soc.}
\newcommand{\apjs}{Astrophys.~J.~Suppl.}
\newcommand{\apjl}{\apj}
\newcommand{\aj}{Astron.~J.}
\newcommand{\jcap}{J.~Cosmol.~Astropart.~Phys.}

@@

\begin{abstract}
We discuss the implications of the Sommerfeld enhancement of the dark
matter annihilation cross section for the detection of dark matter
annihilation from subhalos in the Galactic halo. A robust prediction
of \CDM\ is that the Galactic halo is teeming with cold substructure,
and that the subhalos themselves contain even smaller substructure,
leading to an approximately self-similar distribution of substructure
all the way down to earth-mass halos. In addition to the boost to the
dark matter annihilation cross section from the high densities of
these subhalos, an additional boost due to Sommerfeld enhancement
results from the fact that they are kinematically cold. We take
subhalos from the large numerical $N$-body simulation \VL\ to study
the expected distribution of dark matter annihilation signals from
subhalos at a Sun-like position in a large dark matter halo hosting a
late-type galaxy. We also study the expected Sommerfeld enhanced
annihilation signal for the smallest, most-dark-matter-dominated, and
kinematically coldest dwarf spheroidal satellite galaxy Segue
1. Finally, we discuss the prospects for detecting annihilation
signals from microhalos.
\end{abstract}

@@

\section{Introduction}

First two paragraphs straight from my paper with Glennys:

In the standard flat, Gaussian, adiabatic, and scale-invariant \LCDM\
paradigm $\sim\!80$ \% of the matter density of the Universe is in the
form of dark matter (\DM) \cite{2008arXiv0803.0547K}. The properties
of dark matter are largely unconstrained, although a weakly
interacting massive particle (WIMP) with a mass $\sim\!100$ GeV -- 1
TeV has many attractive features, including that it leads to a relic
density of dark matter that is remarkably close to the measured value
\cite{1996PhR...267..195J}. No interaction of dark matter either with
itself or with standard model (\SM) particles other than gravity has
ever been observed. However, at the present day we cannot exclude that
the dark sector is brimming with a rich phenomenology at
weak-interaction scales or lower.

@@

Of course, the largest obstacle currently to modeling the PAMELA/ATIC
signal with \DM\ annihilation is the very large boost factors ($10^4$
-- $10^5$) with respect to the fiducial $\langle \sigma v
\rangle_{\mathrm{ann}} \approx 3 \times 10^{-26} $ cm$^3$ s$^{-1}$
needed to lead to an observable annihilation flux. The boost factors
from the clumpiness of the dark matter halo is expected to be small
\cite{2008A&A...479..427L,2008Natur.454..735D,2008ApJ...686..262K,2008Natur.456...73S},
typically of the order 1 -- 10. Therefore, a different mechanism must
be reponsible for the large boosts necessary. One such mechanism that
has received much attention lately is the Sommerfeld enhancement
\cite{sommerfeld31a,2003PhRvD..67g5014H,2004PhRvL..92c1303H,2005PhRvD..71f3528H,2005PhRvD..71a5007H,2006PhRvD..73e5004H,2008NuPhB.800..204C,2008JHEP...07..058M,2008arXiv0812.0559M,2008arXiv0812.0360L}. The
Sommerfeld enhancement is a generic effect present whenever there is
an attractive force acting between the dark matter particles, \eg, due
to a Yukawa interaction or a gauge interaction through vector bosons
\cite{ArkaniHamed:2008qn}. The main effect of this attractive force is
to enhance the annihilation cross section with a factor proportional
to $\beta^{-1}$ ($\beta \equiv v/c$), this is known as ``$1/v$''
enhancement. There are regions in the parameter space made up of the
dark matter mass \mdm, the force carrier mass \mv, and the force
``fine-structure constant'' $\alpha \equiv$ coupling$^2$/$4 \pi$ in
which the enhancement is near a resonance, where it approximately
behaves as $\beta^{-2}$. However, the enhancement levels off at a
certain $\beta$ due to the finite range of the attractive force ---
large ($\alpha \sim\!10^{-1}$ -- $10^{-3}$) attractive forces with
infinite range lead to a burst of \DM\ annihilation in the first \DM\
halos formed at $z$ $\sim$ 100 -- 200 which would lead to observable
effects incompatible with measurements of the diffuse extragalactic
gamma-ray background today and the cosmic microwave background
\cite{2008arXiv0810.3233K}.

BOVY: Say something here about clumpiness of the halo: much more than
the following: \cite{1998MNRAS.300..146G,johnston98a,2003MNRAS.339..834H,2008ApJ...688..277T,2002MNRAS.330..778W,2002MNRAS.330..792K,koposov,Koposov:2009ru}

@@

\begin{figure*}[t]
\includegraphics[clip=]{avg-3_64.eps}%%BoundingBox: 178 291 400 474
\includegraphics[clip=]{avg-4_128.eps}\\%%BoundingBox: 190 291 418 474
\includegraphics[clip=]{avg-5_128.eps}%%BoundingBox: 178 291 400 474
\includegraphics[clip=]{avg-6_128.eps}%%BoundingBox: 190 291 418 474
\caption{Sommerfeld enhancement factor as a function of the particle
physics parameters averaged over a Maxwell-Boltzman velocity
distribution with velocity dispersion \sigv\ for four different orders
of magnitude of the velocity dispersion (all velocities are relative
to the speed of light).}%
\label{fig:sommavg}%
\end{figure*}

@@

Further on it will sometimes be useful to describe the \DM\ density
profiles by different parameters than a characteristic density and
radius. Another set of parameters is given by a concentration
parameter and a mass. The concentration for both the \GNFW\ profile as
for the Einasto profile is given by
\begin{equation}\label{eq:concdef}
\conc \equiv \frac{\Rvir}{\rminustwo}\, .
\end{equation}
A characteristic mass is given by the virial mass \Mvir. The virial
mass of the \GNFW\ profile is given by
\begin{equation}\label{eq:Mvirdef}
\Mvir = 4\pi\rhos\rs^3 \left[ - (-\conc)^\gamma\,\conc^{-\gamma} B(-c;3-\gamma,\gamma-2)\right]\, ,
\end{equation}
in which $B(z;a,b)$ is the incomplete Beta function, given by the
familiar form $f(c) = \ln(1+c)-c/(1+c)$ in the special case of the
\NFW\ profile. The virial mass for the Einasto profile is given by
\begin{equation}
\Mvir = 4\pi\rhominustwo\rminustwo^3 \frac{1}{\alphaEinasto} \exp\left(\frac{3\ln \alphaEinasto + 2 - \ln 8}{\alphaEinasto}\right) \gamma\left[\frac{3}{\alphaEinasto},\frac{2}{\alphaEinasto}\conc^\alphaEinasto\right]\, .
\end{equation}
In this last expression $\gamma\left[a,z\right]$ is the lower
incomplete gamma function. The density can then be expressed in terms
of the concentration by using the definition of the virial mass:
$\Mvir \equiv 4\pi/3 \Delta \rhoc \Rvir^3$, where $\Delta = 178\,
\omegam^{0.45} \approx 100$
\cite{1996MNRAS.282..263E,1998ApJ...495...80B}, and \rhoc\ is the
critical density.

Another pair of parameters that one can use to characterize the \DM\
density profile is the maximum circular velocity \vmax\ and the radius
at which this maximum occurs, \rmax. These parameters are especially
well-suited when dealing with numerical simulations, since they can be
readily read off from the raw simulation data. The circular velocity
is given by
\begin{equation}
\vcirc^2(r) \equiv \frac{G M(r)}{r} = \vcirc^2(\Rvir)\frac{c}{f(c)}\frac{f(x)}{x}\, ,
\end{equation}
in which $x \equiv r/\rminustwo$ and $f(x)$ is given by
$B(-x;3-\gamma,\gamma-2)$ in the case of a \GNFW\ profile and by
$\gamma\left[\frac{3}{\alphaEinasto},\frac{2}{\alphaEinasto}x^\alphaEinasto\right]$
in the case of an Einasto profile. One can find \rmax\ by (numerically) solving the equation
\begin{equation}
\frac{\dd}{\dd x}\left[ \frac{f(x)}{x}\right] = 0\, ,
\end{equation}
and \vmax\ follows then immediatly.

@@

In the previous section we calculated $\lumsmoothc$ for the different
\DM\ density profiles and found in all cases that
\begin{equation}
\lumsmoothc(m) \equiv \lumsmoothc(\rho(m),r(m))\, ,
\end{equation}
for a characteristic density $\rho$ and radius $r$, the exact
definition of which depends on the specific \DM\ density profile
used. In order to find the mass dependence of $\lumsmoothc$, we use
the description of the \DM\ density profile in terms of virial mass
\Mvir\ and concentration \conc. We then make use of the fact that many
simulations have found a relation between the concentration of halos
and their mass
\cite{2001MNRAS.321..559B,2001ApJ...554..114E,2008MNRAS.387..536G,2007MNRAS.381.1450N},
such that everything depends only on the mass. This relation is well
established over a large range in halo masses, and we will adopt the
particular relationship found in Ref.~\cite{2007MNRAS.378...55M}
\begin{equation}\label{eq:concscaling}
\conc = 10.5 \, \left(\frac{\Mvir}{10^{12}\, \Msol}\right)^{-0.11}\, ,
\end{equation}
in good agreement with the results of other, similar analyses
\cite{2001MNRAS.321..559B,2005MNRAS.357..387K}, although the
normalization of the concentration-mass relation is about 15\% lower
than the original relation. However, the concentration of subhalos is
generally larger than the concentration of field halos of the same
mass as a result of tidal mass-loss
\cite{1998MNRAS.300..146G,2008ApJ...673..226P,2004MNRAS.355..794H,2005ApJ...635..931B,2004ApJ...608..663K}. This
is a small effect, of order 10\%, and we will investigate later
whether it significantly influences our results.

@@

On the scale of clusters there is a well-defined relationship between
the mass of a \DM\ halo and its one-dimensional velocity dispersion
which one can derive from adiabatic scaling arguments
\cite{1998ApJ...495...80B} given by
\begin{equation}\label{eq:vdispscaling}
\sigma \propto m^{1/3}\, .
\end{equation}
This scaling relation can be easily derived in the case of an
isothermal distribution function \cite{2008gady.book.....B}, since
then
\begin{equation}
\rho(r) = \frac{\sigma^2}{2\pi G r^2}\, ,
\end{equation}
from which \eqnname\ (\ref{eq:vdispscaling}) immediatly follows. This
scaling relation has been found to describe numerical simulations of
cluster-sized halos extremely well
\cite{1991ApJ...383...95E,1995MNRAS.275..720N,1995AJ....110...21C,1996ApJ...469..494E,1996MNRAS.281..716C,1998ApJ...495...80B}.From
recent numerical simulations it seems that we can use this relation
even for subhalos of subhalos in a Galaxy sized dark matter halo. In
Ref.~\cite{2008MNRAS.391.1685S} it is shown that the relation between
the maximum circular velocity of a subhalo and its mass is well fit by
a power-law of the type
\begin{equation}
m \propto \vmax^{3.5}\, .
\end{equation}
By a simple estimate of the velocity dispersion using the virial
theorem, we can use the maximum circular velocity as a measure of the
halos velocity dispersion. This would give $\sigv \propto
m^{1/3.5}$. Alternatively, we could use the circular velocity at the
virial radius as a measure of the velocity dispersion. In the case of
an \NFW\ profile we have that
\begin{equation}
\frac{\vmax^2}{\vcirc(\Rvir)^2} \propto \frac{c}{f(c)}\, ,
\end{equation}
in which $f(x) = \ln(1+c) - c/(1+c) \approx 2.6 (c/33)^{0.4}$ in the
relevant concentration range. Using \eqnname\ (\ref{eq:concscaling}),
this gives
\begin{equation}
\sigv \propto M^{0.32}\, .
\end{equation}
Therefore, we will use (as a rough approximation) the scaling relation
given in \eqnname\ (\ref{eq:vdispscaling}).

\section{A typical distribution of signals in a halo}

\section{Segue 1}

\section{Microhalos}

{\bf Acknowledgments:}
NASA 07-ADP07-0099.

>>>>

\newcommand{\eg}{e.g.}
\newcommand{\Fermi}{\emph{Fermi}}
\newcommand{\VL}{Via Lactea}
\newcommand{\aquarius}{Aquarius}
\newcommand{\CDM}{CDM}
\newcommand{\LCDM}{\ensuremath{\Lambda}\CDM}
\newcommand{\NFW}{NFW}
\newcommand{\GNFW}{G\NFW}
\newcommand{\DM}{DM}
\newcommand{\SM}{SM}
\newcommand{\OM}{OM}
\newcommand{\mpl}{\ensuremath{m_{\mathrm{Pl}}}}
\newcommand{\somm}{\ensuremath{S}}
\newcommand{\mdm}{\ensuremath{m_{\chi}}}
\newcommand{\mv}{\ensuremath{m_{\phi}}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\Eqnname}{Equation}
\newcommand{\eqnname}{equation}
\newcommand{\Ngamma}{\ensuremath{N_{\gamma}}}
\newcommand{\Ngammai}{\ensuremath{N_{\gamma,i}}}
\newcommand{\Eth}{\ensuremath{E_{\mathrm{th}}}}
\newcommand{\sigmaannv}{\ensuremath{\langle\sigma v\rangle}}
\newcommand{\los}{los}
\newcommand{\order}{\ensuremath{\mathcal{O}}}
\newcommand{\sigv}{\ensuremath{\sigma_v}}
\newcommand{\lum}{\ensuremath{\mathcal{L}}}
\newcommand{\lumsmooth}{\ensuremath{\lum_{\mbox{{\footnotesize sm}}}}}
\newcommand{\lumsmoothc}{\ensuremath{\lum_{\mbox{{\footnotesize sm}},v=c}}}
\newcommand{\lumsmoothtoo}{\ensuremath{\overline{\mathcal{L}}}}
\newcommand{\lumsub}{\ensuremath{\lum_{\mbox{{\footnotesize sub}}}}}
\newcommand{\boost}{\ensuremath{B}}
\newcommand{\redr}{\ensuremath{\tilde{r}}}
\newcommand{\rhos}{\ensuremath{\rho_s}}
\newcommand{\rs}{\ensuremath{r_s}}
\newcommand{\dist}{\ensuremath{D}}
\newcommand{\norm}{\ensuremath{\mathcal{N}}}
\newcommand{\normnfw}{\ensuremath{\norm_1}}
\newcommand{\normeinasto}{\ensuremath{\norm_2}}
\newcommand{\rhominustwo}{\ensuremath{\rho_{-2}}}
\newcommand{\rminustwo}{\ensuremath{r_{-2}}}
\newcommand{\sommboost}{\ensuremath{\mathcal{S}}}
\newcommand{\clumpboost}{\ensuremath{\mathcal{C}}}
\newcommand{\alphaEinasto}{\ensuremath{\alpha}}
\newcommand{\Msol}{\ensuremath{M_{\odot}}}
\newcommand{\Msun}{\Msol}
\newcommand{\Mvir}{\ensuremath{M_{\mbox{{\footnotesize vir}}}}}
\newcommand{\Rvir}{\ensuremath{r_{\mbox{{\footnotesize vir}}}}}
\newcommand{\conc}{\ensuremath{c}}
\newcommand{\rhoc}{\ensuremath{\rho_{\mbox{{\footnotesize c}}}}}
\newcommand{\omegam}{\ensuremath{\Omega_{m}}}
\newcommand{\vcirc}{\ensuremath{v_{\mbox{{\footnotesize c}}}}}
\newcommand{\vmax}{\ensuremath{v_{\mbox{{\footnotesize max}}}}}
\newcommand{\rmax}{\ensuremath{r_{\mbox{{\footnotesize max}}}}}

%Journals
\newcommand{\aap}{Astron.~Astrophys.}
\renewcommand{\nat}{Nature}
\newcommand{\jhep}{J.~High.~Energy.~Phys.}
\newcommand{\astropartphys}{Astropart.~Phys.}
\newcommand{\mnras}{Mon.~Not.~Roy.~Astron.~Soc.}
\newcommand{\apjs}{Astrophys.~J.~Suppl.}
\newcommand{\apjl}{\apj}
\newcommand{\aj}{Astron.~J.}
\newcommand{\jcap}{J.~Cosmol.~Astropart.~Phys.}
\newcommand{\araa}{Ann.~Rev.~Astron.~Astrophys.}

@@

\begin{abstract}
I discuss the implications of the recently introduced Sommerfeld
enhancement of the dark matter annihilation cross section for the
detection of subhalos in the Galactic halo.  In addition to the boost
to the dark matter annihilation cross section from the high densities
of these subhalos with respect to the main halo, an additional boost
due to Sommerfeld enhancement results from the fact that they are
kinematically colder than the Galactic halo. If we further believe the
generic prediction of \CDM\ that in each subhalo there is an abundance
of substructure which is approximately self-similar to that of the
Galactic halo, then additional boosts coming from the density
enhancements of these small substructures and their small velocity
dispersions enhance the dark matter annihilation cross section even
further. I show that the main effect of these substructure boosts is
that the Sommerfeld enhancement is pushed to its saturation level in
all the interesting regions of parameters space, and that very large
boost factors ($10^6$ -- $10^9$) are obtained in a large class of
models. The implications of these boost factors for the detection of
dark matter annihilation from dwarf Speroidal galaxies in the Galactic
halo are such that many of the models proposed to explain the positron
excess seen by the PAMELA experiment are already marginally excluded.
\end{abstract}

@@

\section{Introduction}
In the standard flat, Gaussian, adiabatic, and scale-invariant \LCDM\
paradigm $\sim\!80$ \% of the matter density of the Universe is in the
form of dark matter (\DM) \cite{2008arXiv0803.0547K}. The properties
of dark matter are largely unconstrained, although a weakly
interacting massive particle (WIMP) with a mass $\sim\!100$ GeV -- 1
TeV has many attractive features, including that it leads to a relic
density of dark matter that is remarkably close to the measured value
\cite{1996PhR...267..195J}. No interaction of dark matter either with
itself or with standard model (\SM) particles other than gravity has
ever been observed. However, at the present day we cannot exclude that
the dark sector is brimming with a rich phenomenology at
weak-interaction scales or lower.

@@

Of course, the largest obstacle currently to modeling the PAMELA/ATIC
signal with \DM\ annihilation is the very large boost factors ($10^4$
-- $10^5$) with respect to the fiducial $\langle \sigma v
\rangle_{\mathrm{ann}} \approx 3 \times 10^{-26} $ cm$^3$ s$^{-1}$
needed to lead to an observable annihilation flux. The boost factor
from the clumpiness of the dark matter halo is expected to be small
\cite{2008A&A...479..427L,2008Natur.454..735D,2008ApJ...686..262K,2008Natur.456...73S},
typically of the order 1 -- 10. Therefore, a different mechanism must
be reponsible for the large boosts necessary. One such mechanism that
has received much attention lately is the Sommerfeld enhancement
\cite{sommerfeld31a,2003PhRvD..67g5014H,2004PhRvL..92c1303H,2005PhRvD..71f3528H,2005PhRvD..71a5007H,2006PhRvD..73e5004H,2008NuPhB.800..204C,2008JHEP...07..058M,2008arXiv0812.0559M,2008arXiv0812.0360L}. The
Sommerfeld enhancement is a generic effect present whenever there is
an attractive force acting between the dark matter particles, \eg, due
to a Yukawa interaction or a gauge interaction through vector bosons
\cite{ArkaniHamed:2008qn}. The main effect of this attractive force is
to enhance the annihilation cross section with a factor proportional
to $\beta^{-1}$ ($\beta \equiv v/c$), this is known as ``$1/v$''
enhancement. There are regions in the parameter space made up of the
dark matter mass \mdm, the force carrier mass \mv, and the force
``fine-structure constant'' $\alpha \equiv$ coupling$^2$/$4 \pi$ in
which the enhancement is near a resonance, where it approximately
behaves as $\beta^{-2}$. However, the enhancement levels off at a
certain $\beta$ due to the finite range of the attractive force ---
large ($\alpha \sim\!10^{-1}$ -- $10^{-3}$) attractive forces with
infinite range lead to a burst of \DM\ annihilation in the first \DM\
halos formed at $z$ $\sim$ 100 -- 200 which would lead to observable
effects incompatible with measurements of the diffuse extragalactic
gamma-ray background today and the cosmic microwave background
\cite{2008arXiv0810.3233K}. Since the typical velocities of dark
matter particles in the Galactic halo today are of the order of the
velocity dispersion of the halo, which is non-relativistic ($\beta\sim
10^{-3}$), the cross section of dark matter particles annihilating in
the halo could be significantly enhanced by this Sommerfeld enhancement.

Both observations
\cite{1994Natur.370..194I,1999Natur.402...53H,2005ApJ...626L..85W,2007ApJ...654..897B,koposov}
as well as numerical simulations
\cite{1998MNRAS.300..146G,johnston98a,1999ApJ...524L..19M,1999ApJ...522...82K,1999MNRAS.307..495H,2003MNRAS.339..834H,2008Natur.454..735D,2008MNRAS.391.1685S}
have firmly established that dark matter halos in \CDM\ are not smooth
structures, but that they are filled with a large amount of
substructure. These substructures are both in the form of subhalos,
self-bound objects orbiting in the Galactic halo, as well as in
streams, relics of tidally disrupted subhalos. Since these
substructures are overdense regions in the halo, the dark matter
annihilation cross section will be correspondingly higher, since
annihilation cross sections scale as the density squared. The effects
of these kinds of substructure boosts have been studied extensively
during the last decade
\cite{2004PhRvD..70b3512B,2004PhRvD..69d3501K,2007ApJ...657..262D,2007PhRvD..75h3526S,2008A&A...479..427L,2008MNRAS.384.1627P,2008ApJ...678..614S,2008ApJ...686..262K,2008Natur.456...73S,2008arXiv0809.2781G}. However,
these subhalos are not only denser than the Galactic halo in which
they orbit, they are also kinematically colder than the Galactic halo,
with velocity dispersion typically being two orders of magnitude
smaller \cite{1998ARA&A..36..435M,2007ApJ...670..313S}. This means
that the Sommerfeld enhancement in these subhalos can be many orders
of magnitude larger than the Sommerfeld enhancement acting in the
Galactic halo.

In addition to this larger boost due to the small
velocity dispersion in the subhalos, these subhalos themselves are
filled with an abundance of substructure which is approximately
self-similar to that of the main halo. Each of these even smaller
substructure has a larger density than the subhalo they are orbiting
in (quantified by the well established concentration-mass
relationships for \CDM\ halos
\cite{2001MNRAS.321..559B,2001ApJ...554..114E}) as well as a smaller
velocity dispersion. This opens up the possibility that the dark
matter annihilation signal in these already cold subhalos can be
boosted even further by contribution of their substructure. In this
paper we quantify the magnitude of this new substructure boost and
discuss its implications for the detection of dark matter annihilation
from subhalos in the Galactic halos.

@@

\begin{figure*}[t]
\includegraphics[clip=]{avg-3_65.eps}%%BoundingBox: 178 291 400 474
\includegraphics[clip=]{avg-4_129.eps}\\%%BoundingBox: 190 291 418 474
\includegraphics[clip=]{avg-5_129.eps}%%BoundingBox: 178 291 400 474
\includegraphics[clip=]{avg-6_129.eps}%%BoundingBox: 190 291 418 474
\caption{Sommerfeld enhancement factor as a function of the particle
physics parameters averaged over a Maxwell-Boltzman velocity
distribution with velocity dispersion \sigv\ for four different orders
of magnitude of the velocity dispersion (all velocities are relative
to the speed of light).}%
\label{fig:sommavg}%
\end{figure*}

@@

Further on it will sometimes be useful to describe the \DM\ density
profiles by different parameters than a characteristic density and
radius. Another set of parameters is given by a concentration
parameter and a mass. The concentration for both the \GNFW\ profile as
for the Einasto profile is given by
\begin{equation}\label{eq:concdef}
\conc \equiv \frac{\Rvir}{\rminustwo}\, .
\end{equation}
A characteristic mass is given by the virial mass \Mvir. The virial
mass of the \GNFW\ profile is given by
\begin{equation}\label{eq:Mvirdef}
\Mvir = 4\pi\rhos\rs^3 f(c)\, ,
\end{equation}
in which the function $f(c)$ is given by
\begin{equation}
f(c) = \frac{c^{3-\gamma}}{3-\gamma} \, {_2F_1}\left(3-\gamma,3-\gamma;4-\gamma;-c\right)\, .
\end{equation}\label{eq:hypergeometric}
The function $_2 F_1\left(a,b;c;z\right)$ is the Gauss hypergeometric
function. In the $\gamma = 1$ case $f(c)$ is given by the familiar
form $f(c) = \ln(1+c)-c/(1+c)$. The virial mass for the Einasto
profile is given by
\begin{equation}
\Mvir = 4\pi\rhominustwo\rminustwo^3 \frac{1}{\alphaEinasto} \exp\left(\frac{3\ln \alphaEinasto + 2 - \ln 8}{\alphaEinasto}\right) \gamma\left[\frac{3}{\alphaEinasto},\frac{2}{\alphaEinasto}\conc^\alphaEinasto\right]\, .
\end{equation}
In this last expression $\gamma\left[a,z\right]$ is the lower
incomplete gamma function. The density can then be expressed in terms
of the concentration by using the definition of the virial mass:
$\Mvir \equiv 4\pi/3 \Delta \rhoc \Rvir^3$, where $\Delta = 178\,
\omegam^{0.45} \approx 100$
\cite{1996MNRAS.282..263E,1998ApJ...495...80B}, and \rhoc\ is the
critical density.

Another pair of parameters that one can use to characterize the \DM\
density profile is the maximum circular velocity \vmax\ and the radius
at which this maximum occurs, \rmax. These parameters are especially
well-suited when dealing with numerical simulations, since they can be
readily read off from the raw simulation data. The circular velocity
is given by
\begin{equation}
\vcirc^2(r) \equiv \frac{G M(r)}{r} = \vcirc^2(\Rvir)\frac{c}{f(c)}\frac{f(x)}{x}\, ,
\end{equation}
in which $x \equiv r/\rminustwo$ and $f(x)$ is defined in \eqnname\
(\ref{eq:hypergeometric}) in the case of a \GNFW\ profile and is given
by
$\gamma\left[\frac{3}{\alphaEinasto},\frac{2}{\alphaEinasto}x^\alphaEinasto\right]$
in the case of an Einasto profile. One can find \rmax\ by
(numerically) solving the equation
\begin{equation}
\frac{\dd}{\dd x}\left[ \frac{f(x)}{x}\right] = 0\, ,
\end{equation}
and \vmax\ follows then immediatly.

@@

In the previous section we calculated $\lumsmoothc$ for the different
\DM\ density profiles and found in all cases that
\begin{equation}
\lumsmoothc(m) \equiv \lumsmoothc(\rho(m),r(m))\, ,
\end{equation}
for a characteristic density $\rho$ and radius $r$, the exact
definition of which depends on the specific \DM\ density profile
used. In order to find the mass dependence of $\lumsmoothc$, we use
the description of the \DM\ density profile in terms of virial mass
\Mvir\ and concentration \conc. We then make use of the fact that many
simulations have found a relation between the concentration of halos
and their mass
\cite{2001MNRAS.321..559B,2001ApJ...554..114E,2008MNRAS.387..536G,2007MNRAS.381.1450N},
such that everything depends only on the mass. This relation is well
established over a large range in halo masses, and we will adopt the
particular relationship found in Ref.~\cite{2007MNRAS.378...55M}
\begin{equation}\label{eq:concscaling}
\conc = 10.5 \, \left(\frac{\Mvir}{10^{12}\, \Msol}\right)^{-0.11}\, ,
\end{equation}
in good agreement with the results of other, similar analyses
\cite{2001MNRAS.321..559B,2005MNRAS.357..387K}, although the
normalization of the concentration-mass relation is about 15\% lower
than the original relation. However, the concentration of subhalos is
generally larger than the concentration of field halos of the same
mass as a result of tidal mass-loss
\cite{1998MNRAS.300..146G,2008ApJ...673..226P,2004MNRAS.355..794H,2005ApJ...635..931B,2004ApJ...608..663K}. This
is a small effect, of order 10\%, and it does not play a large role in
the results we obtain below. This concentration-halo mass
concentration has not just been establised in numerical simulations:
statistical weak lensing analysis of luminous red galaxies and
clusters from the Sloan Digital Sky Survey has shown that the halo
profiles of these objects are consistent with the \NFW\ profile and
mass-concentration relation given above \cite{2008JCAP...08..006M}.

@@

On the scale of clusters there is a well-defined relationship between
the mass of a \DM\ halo and its one-dimensional velocity dispersion
which one can derive from adiabatic scaling arguments
\cite{1998ApJ...495...80B} given by
\begin{equation}\label{eq:vdispscaling}
\sigma \propto m^{1/3}\, .
\end{equation}
This scaling relation can be easily derived in the case of an
isothermal distribution function \cite{2008gady.book.....B}, since
then
\begin{equation}
\rho(r) = \frac{\sigma^2}{2\pi G r^2}\, ,
\end{equation}
from which \eqnname\ (\ref{eq:vdispscaling}) immediatly follows. This
scaling relation has been found to describe numerical simulations of
cluster-sized halos extremely well
\cite{1991ApJ...383...95E,1995MNRAS.275..720N,1995AJ....110...21C,1996ApJ...469..494E,1996MNRAS.281..716C,1998ApJ...495...80B}.From
recent numerical simulations it seems that we can use this relation
even for subhalos of subhalos in a Galaxy sized dark matter halo. In
Ref.~\cite{2008MNRAS.391.1685S} it is shown that the relation between
the maximum circular velocity of a subhalo and its mass is well fit by
a power-law of the type
\begin{equation}
m \propto \vmax^{3.5}\, .
\end{equation}
By a simple estimate of the velocity dispersion using the virial
theorem, we can use the maximum circular velocity as a measure of the
halos velocity dispersion. This would give $\sigv \propto
m^{1/3.5}$. Alternatively, we could use the circular velocity at the
virial radius as a measure of the velocity dispersion. In the case of
an \NFW\ profile we have that
\begin{equation}
\frac{\vmax^2}{\vcirc(\Rvir)^2} \propto \frac{c}{f(c)}\, ,
\end{equation}
in which $f(x) = \ln(1+c) - c/(1+c) \approx 2.6 (c/33)^{0.4}$ in the
relevant concentration range. Using \eqnname\ (\ref{eq:concscaling}),
this gives
\begin{equation}
\sigv \propto M^{0.32}\, .
\end{equation}
Therefore, we will use (as a rough approximation) the scaling relation
given in \eqnname\ (\ref{eq:vdispscaling}). We calibrate this
relationship at the mass scale of the dwarf spheroidals by using the
velocity and mass of Canes Venatici I from Ref.~\cite{2007ApJ...670..313S}.

\subsection{The total and the ``reduced'' boost}

\begin{figure*}[t]
\includegraphics[clip=]{2dboost8_65.eps}%%BoundingBox: 178 291 400 474
\includegraphics[clip=]{2dboost7_65.eps}\\%%BoundingBox: 190 291 430 474
\includegraphics[clip=]{2dredboost8_65.eps}%%BoundingBox: 170 277 398 487
\includegraphics[clip=]{2dredboost7_65.eps}%%BoundingBox: 192 277 432 487
\caption{}%
\label{fig:subboost}%
\end{figure*}

\begin{figure*}[t]
\includegraphics[clip=]{2dboost8_65_0.8.eps}%%BoundingBox: 178 291 400 474
\includegraphics[clip=]{2dboost8_65_1.4.eps}\\%%BoundingBox: 190 291 430 474
%\includegraphics[clip=]{2dredboost8_65.eps}%%BoundingBox: 178 291 400 474
%\includegraphics[clip=]{2dredboost7_65.eps}%%BoundingBox: 190 291 430 474
\caption{}%
\label{fig:subboostprofiles}%
\end{figure*}

\section{Predicted annihilation signals for dwarf Spheroidals}
